---
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    df_print: paged
    theme: cosmo
    toc: true
    toc_float: true
---


```{r snakemake_object, include=FALSE, eval=FALSE}
if (interactive()) { ## fake data for testing
  Snakemake <- setClass("Snakemake", slots=c(input="list", output="list",
                                             params="list", threads="numeric"))
  snakemake <- Snakemake(
    input=list(bedpe_1="/work/aaa/projects/chrombpnet-devmult/pipeline/results/chrombpnet_nobias/pretrained_bias/Large-PreB/mean/modisco/Large-PreB_mean.counts_scores.top_tf.bedpe.gz",
               bedpe_2="/work/aaa/projects/chrombpnet-devmult/pipeline/results/chrombpnet_nobias/pretrained_bias/IM-B/mean/modisco/IM-B_mean.counts_scores.top_tf.bedpe.gz"),
    output=list("/work/aaa/projects/chrombpnet-devmult/pipeline/results/reports/seqlet-overlap/PreProB_IM-B.mean.counts_counts.html"),
    params=list(group_1="Large-PreB",
                group_2="IM-B"),
    threads=8L)
}

print(snakemake)
```


```{r echo=FALSE}
title_str <- sprintf("ChromBPNet Seqlet Overlaps - %s vs %s", snakemake@params$group_1, snakemake@params$group_2)
```

---
title: `r title_str`
---


# Purpose

Here we contrast the ChromBPNet seqlets across **`r snakemake@params$group_1`** and **`r snakemake@params$group_2`**.
We are interested in both how much is shared and how much is unique.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(grid)
library(tidyverse)
library(GenomicRanges)
library(plyranges)
library(BiocParallel)
library(pheatmap)
```

## Parameters
```{r set_params}
set.seed(20250508)

register(MulticoreParam(workers=snakemake@threads))
theme_set(theme_bw())
```

## Functions

```{r methods}
read_bedpe_seqlet <- function (file) {
  read_tsv(file,
           col_names=c("seqnames", "start", "end",
                       "chrom_peak", "start_peak", "end_peak",
                       "name", "score", "strand", "strand_peak",
                       "score_target", "pattern_id"),
           col_types="ciiciicdffdc") %>%
    as_granges()
}

read_bedpe_peak <- function (file) {
  read_tsv(file,
           col_names=c("chrom_seqlet", "start_seqlet", "end_seqlet",
                       "seqnames", "start", "end",
                       "name", "score_seqlet", "strand_seqlet", "strand",
                       "score", "pattern_id"),
           col_types="ciiciicdffdc") %>%
    as_granges()
}

# Jaccard function
jaccard_index <- function(gr_a, gr_b) {
  intersected <- sum(width(reduce(intersect(gr_a, gr_b))))
  unioned     <- sum(width(reduce(union(gr_a, gr_b))))
  if (unioned == 0) return(NA)  # avoid division by zero
  intersected / unioned
}

# Overlap index function
overlap_index <- function(gr_a, gr_b) {
  intersected <- sum(width(reduce(intersect(gr_a, gr_b))))
  min_total   <- min(sum(width(reduce(gr_a))), sum(width(reduce(gr_b))))
  if (min_total == 0) return(NA)
  intersected / min_total
}

# Overlap index function
overlap_index_left <- function(gr_a, gr_b) {
  intersected <- sum(count_overlaps(gr_a, gr_b) > 0)
  total_left <- length(gr_a)
  if (total_left == 0) return(NA)
  intersected / total_left
}
```

# Data
## Loading
```{r load_data, message=FALSE}
gr_seqlets1 <- read_bedpe_seqlet(snakemake@input$bedpe_1)
gr_seqlets2 <- read_bedpe_seqlet(snakemake@input$bedpe_2)
gr_peaks1 <- read_bedpe_peak(snakemake@input$bedpe_1)
gr_peaks2 <- read_bedpe_peak(snakemake@input$bedpe_2)
```

# Overview

## Peak Counts

```{r plt_peak_counts, fig.width=5, fig.height=8}
n_peaks1 <- gr_peaks1 %>%
  as_data_frame() %>%
  distinct(seqnames, start, end) %>%
  nrow()

n_peaks2 <- gr_peaks2 %>%
  as_data_frame() %>%
  distinct(seqnames, start, end) %>%
  nrow()

df_npeaks1 <- gr_peaks1 %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  as_data_frame() %>%
  distinct(seqnames, start, end, pattern_group, name) %>%
  count(pattern_group, name, name="n_peaks")

df_npeaks2 <- gr_peaks2 %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  as_data_frame() %>%
  distinct(seqnames, start, end, pattern_group, name) %>%
  count(pattern_group, name, name="n_peaks")

df_npeaks1 %>%
  ggplot(aes(x=n_peaks, y=fct_infreq(pattern_group, n_peaks))) +
  geom_bar(stat='identity', fill="steelblue2", color="black", linewidth=0.2) +
  scale_x_log10(expand=c(0,0,0.05,0)) +
  labs(title=snakemake@params$group_1, x="Peaks", y=NULL)

df_npeaks2 %>%
  ggplot(aes(x=n_peaks, y=fct_infreq(pattern_group, n_peaks))) +
  geom_bar(stat='identity', fill="steelblue2", color="black", linewidth=0.2) +
  scale_x_log10(expand=c(0,0,0.05,0)) +
  labs(title=snakemake@params$group_2, x="Peaks", y=NULL)
```

In total there are `r n_peaks1` peaks in `r snakemake@params$group_1` and `r n_peaks2` peaks in `r snakemake@params$group_2`.

## Seqlet Counts

There are `r length(gr_seqlets1)` seqlets in `r snakemake@params$group_1` and `r length(gr_seqlets2)` seqlets in `r snakemake@params$group_2`.

```{r plt_nseqlets, fig.width=5, fig.height=8}
df_nseqlets1 <- gr_seqlets1 %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  as_data_frame() %>%
  distinct(seqnames, start, end, pattern_group, name) %>%
  count(pattern_group, name, name="n_seqlets")
df_nseqlets2 <- gr_seqlets2 %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  as_data_frame() %>%
  distinct(seqnames, start, end, pattern_group, name) %>%
  count(pattern_group, name, name="n_seqlets")


df_nseqlets1 %>%
  ggplot(aes(x=n_seqlets, y=fct_infreq(pattern_group, n_seqlets))) +
  geom_bar(stat='identity', fill="steelblue2", color="black", linewidth=0.2) +
  scale_x_log10(expand=c(0,0,0.05,0)) +
  labs(title=snakemake@params$group_1, x="Seqlets", y=NULL)

df_nseqlets2 %>%
  ggplot(aes(x=n_seqlets, y=fct_infreq(pattern_group, n_seqlets))) +
  geom_bar(stat='identity', fill="steelblue2", color="black", linewidth=0.2) +
  scale_x_log10(expand=c(0,0,0.05,0)) +
  labs(title=snakemake@params$group_2, x="Seqlets", y=NULL)
```
## Seqlets Per Peak
Are there some seqlets that tend to occur repeatedly within a peak?

```{r plt_seq_per_pk, fig.width=3, fig.height=8}
full_join(df_npeaks1, df_nseqlets1, by=c("pattern_group", "name")) %>%
  mutate(n_seq_per_peak=n_seqlets/n_peaks) %>%
  ggplot(aes(x=n_seq_per_peak, y=fct_infreq(pattern_group, n_seq_per_peak))) +
  geom_point() + geom_vline(xintercept=1, linetype="dashed", color="orchid", linewidth=0.5) +
  labs(title=snakemake@params$group_1, x="Seqlets per Peak", y=NULL)
  
full_join(df_npeaks2, df_nseqlets2, by=c("pattern_group", "name")) %>%
  mutate(n_seq_per_peak=n_seqlets/n_peaks) %>%
  ggplot(aes(x=n_seq_per_peak, y=fct_infreq(pattern_group, n_seq_per_peak))) +
  geom_point() + geom_vline(xintercept=1, linetype="dashed", color="orchid", linewidth=0.5) +
  labs(title=snakemake@params$group_2, x="Seqlets per Peak", y=NULL)
```

# Comparison
## Total Peak Overlap
How many peaks overlap in total?

```{r plt_pk_overlap_total, fig.width=3, fig.height=3}
gr_peaks_uniq1 <- gr_peaks1 %>%
  as_data_frame() %>%
  distinct(seqnames, start, end) %>%
  as_granges()

gr_peaks_uniq2 <- gr_peaks2 %>%
  as_data_frame() %>%
  distinct(seqnames, start, end) %>%
  as_granges()

n_overlapping1 <-sum(count_overlaps(gr_peaks_uniq1, gr_peaks_uniq2) > 0)
n_overlapping2 <-sum(count_overlaps(gr_peaks_uniq2, gr_peaks_uniq1) > 0)

tibble(group_id=c(snakemake@params$group_1, snakemake@params$group_1,
                  snakemake@params$group_2, snakemake@params$group_2),
       peak_type=fct(c("shared", "unique", "shared", "unique"), levels=c("unique", "shared")),
       n_peaks=c(n_overlapping1, n_peaks1 - n_overlapping1,
                 n_overlapping2, n_peaks2 - n_overlapping2)) %>%
  ggplot(aes(x=group_id, y=n_peaks, fill=peak_type)) +
  geom_bar(stat='identity', position='stack', color='black', linewidth=0.3) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  scale_fill_manual(values=c("unique"="grey80", shared="steelblue3")) +
  labs(x=NULL, y="Peaks", fill="Type")
```

## Total Seqlet Overlap
How many seqlets overlap in total?

```{r plt_seq_overlap_total, fig.width=3, fig.height=3}
n_overlapping1 <-sum(count_overlaps(gr_seqlets1, gr_seqlets2) > 0)
n_overlapping2 <-sum(count_overlaps(gr_seqlets2, gr_seqlets1) > 0)

tibble(group_id=c(snakemake@params$group_1, snakemake@params$group_1,
                  snakemake@params$group_2, snakemake@params$group_2),
       peak_type=fct(c("shared", "unique", "shared", "unique"), levels=c("unique", "shared")),
       n_seqlets=c(n_overlapping1, length(gr_seqlets1) - n_overlapping1,
                 n_overlapping2, length(gr_seqlets2) - n_overlapping2)) %>%
  ggplot(aes(x=group_id, y=n_seqlets, fill=peak_type)) +
  geom_bar(stat='identity', position='stack', color='black', linewidth=0.3) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  scale_fill_manual(values=c("unique"="grey80", shared="steelblue3")) +
  labs(x=NULL, y="Seqlets", fill="Type")
```

## Seqlet Overlap by Pattern
### Jaccard Index
The Jaccard index checks the fraction of shared among the union between pairs of patterns. This indicates the total
amount of sharing between patterns.

```{r compute_jaccard, fig.width=12, fig.height=12}
# Split by pattern_group
gr1_split <- gr_seqlets1 %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  split(f=.$pattern_group)
gr2_split <- gr_seqlets2 %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  split(f=.$pattern_group)

idx1 <- names(gr1_split)
idx2 <- names(gr2_split)

combo_df <- expand.grid(pattern1=idx1, pattern2=idx2, stringsAsFactors = FALSE)

# compute jaccard
jac_values <- bplapply(seq_len(nrow(combo_df)), function(i) {
  row <- combo_df[i, ]
  jaccard_index(gr1_split[[row$pattern1]], gr2_split[[row$pattern2]])
})

mat_jac <- matrix(unlist(jac_values), nrow = length(idx1), ncol = length(idx2),
                  dimnames = list(idx1, idx2), byrow = FALSE)
```


#### Full Scale
```{r plt_jaccard, fig.width=12, fig.height=12}
pheatmap(mat_jac, cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```


#### Threshold (Max 0.1)
```{r plt_jaccard2, fig.width=12, fig.height=12}
pheatmap(mat_jac, cluster_rows=TRUE, cluster_cols=TRUE, breaks = seq(0, 0.1, length.out = 100))
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```

### Overlap Index
The overlap index checks the fraction of shared within the smallest set. This identifies subset-superset relationships.

```{r compute_overlapi, fig.width=12, fig.height=12}
overlap_values <- bplapply(seq_len(nrow(combo_df)), function(i) {
  row <- combo_df[i, ]
  overlap_index(gr1_split[[row$pattern1]], gr2_split[[row$pattern2]])
})

mat_overlap <- matrix(unlist(overlap_values), nrow=length(idx1), ncol=length(idx2),
                      dimnames=list(idx1,idx2), byrow=FALSE)
```


#### Full Scale
```{r plt_overlapi, fig.width=12, fig.height=12}
pheatmap(mat_overlap, cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```


#### Threshold (Max 0.1)
```{r plt_overlapi2, fig.width=12, fig.height=12}
pheatmap(mat_overlap, cluster_rows=TRUE, cluster_cols=TRUE, breaks=seq(0,0.1,length.out=100))
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```


### Rare Overlaps
#### `r snakemake@params$group_1` Seqlets
```{r plt_overlapi_rare_row, fig.width=6, fig.height=6}
idx_row <- rowSums(mat_overlap >= 0.5) == 0
idx_col <- colSums(mat_overlap[idx_row,] >= 0.05) > 0
pheatmap(mat_overlap[idx_row, idx_col], cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```

#### `r snakemake@params$group_2` Seqlets
```{r plt_overlapi_rare_col, fig.width=6, fig.height=6}
idx_col <- colSums(mat_overlap >= 0.5) == 0
idx_row <- rowSums(mat_overlap[,idx_col] >= 0.05) > 0
pheatmap(mat_overlap[idx_row, idx_col], cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```

### Directed Overlap Index
The directed overlap index checks the fraction of shared among the first set. This identifies subset-superset relationships.

#### Fraction of `r snakemake@params$group_1` Seqlets
```{r compute_doverlapi1, fig.width=12, fig.height=12}
doverlap_values1 <- bplapply(seq_len(nrow(combo_df)), function(i) {
  row <- combo_df[i, ]
  overlap_index_left(gr1_split[[row$pattern1]], gr2_split[[row$pattern2]])
})

mat_doverlap1 <- matrix(unlist(doverlap_values1), nrow=length(idx1), ncol=length(idx2),
                      dimnames=list(idx1,idx2), byrow=FALSE)
```


```{r plt_doverlapi1, fig.width=12, fig.height=12}
pheatmap(mat_doverlap1, cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```


#### Rare Overlaps
```{r plt_doverlapi_rare, fig.width=8, fig.height=8}
idx_row <- rowSums(mat_doverlap1 >= 0.5) == 0
idx_col <- colSums(mat_doverlap1[idx_row,] >= 0.05) > 0
pheatmap(mat_doverlap1[idx_row, idx_col], cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```

#### Fraction of `r snakemake@params$group_2` Seqlets
```{r compute_doverlapi2, fig.width=12, fig.height=12}
doverlap_values2 <- bplapply(seq_len(nrow(combo_df)), function(i) {
  row <- combo_df[i, ]
  overlap_index_left(gr2_split[[row$pattern2]], gr1_split[[row$pattern1]])
})

mat_doverlap2 <- matrix(unlist(doverlap_values2), nrow=length(idx1), ncol=length(idx2),
                      dimnames=list(idx1,idx2), byrow=FALSE)
```


```{r plt_doverlapi2, fig.width=12, fig.height=12}
pheatmap(mat_doverlap2, cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```


#### Rare Overlaps
```{r plt_doverlapi_rare2, fig.width=8, fig.height=8}
idx_col <- colSums(mat_doverlap2 >= 0.5) == 0
idx_row <- rowSums(mat_doverlap2[,idx_col] >= 0.05) > 0
pheatmap(mat_doverlap2[idx_row, idx_col], cluster_rows=TRUE, cluster_cols=TRUE)
# x-axis label
grid.text(snakemake@params$group_2, y=unit(0.98, "npc"))
# y-axis label
grid.text(snakemake@params$group_1, x=unit(0.02, "npc"), rot=90)
```


---

# Runtime Details
## Session Info
<details>
```{r sesh_info, echo=FALSE}
sessionInfo()
```
</details>

## Conda Environment
<details>
```{bash conda_info, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  PREFIX=$(dirname $(dirname $(which R)))
  conda env export -p $PREFIX
fi
```
</details>
