---
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    df_print: paged
    theme: cosmo
    toc: true
    toc_float: true
---


```{r snakemake_object, include=FALSE, eval=FALSE}
if (interactive()) { ## fake data for testing
  Snakemake <- setClass("Snakemake", slots=c(input="list", output="list",
                                             params="list", threads="numeric"))
  snakemake <- Snakemake(
    input=list(bed_pk="/work/aaa/projects/chrombpnet-devmult/pipeline/resources/peaks/IM-B/IM-B.no_blacklist.bed",
               bedpe="/work/aaa/projects/chrombpnet-devmult/pipeline/results/chrombpnet_nobias/pretrained_bias/IM-B/mean/modisco/IM-B_mean.counts_scores.top_tf.bedpe.gz"),
    output=list("/work/aaa/projects/chrombpnet-devmult/pipeline/results/reports/summarize_seqlets/mean/report_seqlet_summary.IM-B_mean.counts_scores.html"),
    params=list(sample="Large-PreB"),
    threads=8L)
}

print(snakemake)
```


```{r echo=FALSE}
title_str <- sprintf("ChromBPNet Seqlet Summary - %s", snakemake@params$sample)
```

---
title: `r title_str`
---


# Purpose

Here we summarize the ChromBPNet seqlets in the **`r snakemake@params$sample`** sample.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(grid)
library(tidyverse)
library(broom)
library(GenomicRanges)
library(plyranges)
library(BiocParallel)
library(pheatmap)
```

## Parameters
```{r set_params}
set.seed(20250508)

register(MulticoreParam(workers=snakemake@threads))
theme_set(theme_bw())
```

## Functions

```{r methods}
read_bedpe_seqlet <- function (file) {
  read_tsv(file,
           col_names=c("seqnames", "start", "end",
                       "chrom_peak", "start_peak", "end_peak",
                       "name", "score", "strand", "strand_peak",
                       "score_target", "pattern_id"),
           col_types="ciiciicdffdc") %>%
    as_granges()
}

read_bedpe_peak <- function (file) {
  read_tsv(file,
           col_names=c("chrom_seqlet", "start_seqlet", "end_seqlet",
                       "seqnames", "start", "end",
                       "name", "score_seqlet", "strand_seqlet", "strand",
                       "score", "pattern_id"),
           col_types="ciiciicdffdc") %>%
    as_granges()
}

# Jaccard function
jaccard_index <- function(gr_a, gr_b) {
  intersected <- sum(width(reduce(intersect(gr_a, gr_b))))
  unioned     <- sum(width(reduce(union(gr_a, gr_b))))
  if (unioned == 0) return(NA)  # avoid division by zero
  intersected / unioned
}

# Overlap index function
overlap_index <- function(gr_a, gr_b) {
  intersected <- sum(width(reduce(intersect(gr_a, gr_b))))
  min_total   <- min(sum(width(reduce(gr_a))), sum(width(reduce(gr_b))))
  if (min_total == 0) return(NA)
  intersected / min_total
}

# Overlap index function
overlap_index_left <- function(gr_a, gr_b) {
  intersected <- sum(count_overlaps(gr_a, gr_b) > 0)
  total_left <- length(gr_a)
  if (total_left == 0) return(NA)
  intersected / total_left
}
```

# Data
## Loading
```{r load_data, message=FALSE}
gr_seqlets <- read_bedpe_seqlet(snakemake@input$bedpe)
gr_peaks <- read_bedpe_peak(snakemake@input$bedpe)

n_total_raw <- length(read_lines(snakemake@input$bed_pk))
```

# Overview

## Peak Counts

```{r plt_peak_counts, fig.width=5, fig.height=8}
n_peaks <- gr_peaks %>%
  as_data_frame() %>%
  distinct(seqnames, start, end) %>%
  nrow()

df_npeaks <- gr_peaks %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  as_data_frame() %>%
  distinct(seqnames, start, end, pattern_group, name) %>%
  count(pattern_group, name, name="n_peaks")

df_npeaks %>%
  ggplot(aes(x=n_peaks, y=fct_infreq(pattern_group, n_peaks))) +
  geom_bar(stat='identity', fill="steelblue2", color="black", linewidth=0.2) +
  scale_x_log10(expand=c(0,0,0.05,0)) +
  labs(x="Peaks", y=NULL)
```

In total there were `r n_total_raw` input peaks in `r snakemake@params$sample`. Among these, `r n_peaks` (`r 100*n_peaks/n_total_raw`%%) had seqlets detected.

## Seqlet Counts

There are `r length(gr_seqlets)` seqlets in `r snakemake@params$sample`.

```{r plt_nseqlets, fig.width=5, fig.height=8}
df_nseqlets <- gr_seqlets %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns.")) %>%
  as_data_frame() %>%
  distinct(seqnames, start, end, pattern_group, name) %>%
  count(pattern_group, name, name="n_seqlets")

df_nseqlets %>%
  ggplot(aes(x=n_seqlets, y=fct_infreq(pattern_group, n_seqlets))) +
  geom_bar(stat='identity', fill="steelblue2", color="black", linewidth=0.2) +
  scale_x_log10(expand=c(0,0,0.05,0)) +
  labs(x="Seqlets", y=NULL)
```

## Seqlets Per Peak
On average, how many seqlets are in each peak?

```{r plt_pk_overlap_all, fig.width=3.5, fig.height=3}
gr_seqlets %>%
  as_tibble() %>%
  count(chrom_peak, start_peak, end_peak, name="n_seqlets") %>%
  mutate(n_seqlets_fct=fct(case_when(n_seqlets >= 4 ~ "4+", TRUE ~ as.character(n_seqlets)),
                           levels=c(0,1,2,3,"4+"))) %>%
  count(n_seqlets_fct, name="n_peaks") %>% 
  bind_rows(tibble(n_seqlets_fct="0", n_peaks=n_total_raw - n_peaks)) %>%
  ggplot(aes(x=n_seqlets_fct, y=n_peaks)) +
  geom_bar(stat='identity', fill='steelblue', color='black', linewidth=0.2) +
  scale_x_discrete(drop=FALSE) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x="Seqlets per peak", y="Peaks")
```


```{r plt_pk_overlap_hits, fig.width=3, fig.height=3}
gr_seqlets %>%
  as_tibble() %>%
  count(chrom_peak, start_peak, end_peak, name="n_seqlets") %>%
  mutate(n_seqlets_fct=fct(case_when(n_seqlets >= 4 ~ "4+", TRUE ~ as.character(n_seqlets)),
                           levels=c(1,2,3,"4+"))) %>%
  ggplot(aes(x=n_seqlets_fct)) +
  geom_bar(fill='steelblue', color='black', linewidth=0.2) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x="Seqlets per peak", y="Peaks")
```


## Repeating Seqlets
Are there some seqlets that tend to occur repeatedly within a peak?

```{r plt_seq_per_pk, fig.width=3, fig.height=8}
full_join(df_npeaks, df_nseqlets, by=c("pattern_group", "name")) %>%
  mutate(n_seq_per_peak=n_seqlets/n_peaks) %>%
  ggplot(aes(x=n_seq_per_peak, y=fct_infreq(pattern_group, n_seq_per_peak))) +
  geom_point() + geom_vline(xintercept=1, linetype="dashed", color="orchid", linewidth=0.5) +
  labs(x="Seqlets per Peak", y=NULL)
```

## Frequent Combinations
Are there some combinations that frequently occur together?

First, we will compute the matrix of frequencies. This will have conditional counts based on peaks with a given TFF.
That is, each entry will record the number of times the column TFF occurred in a peak with the row TFF. Accordingly,
the diagonal will correspond to the total number of peaks that a TFF occurred.

```{r compute_conditional_pair_freqs}
df_seqlets <- gr_seqlets %>%
  mutate(pattern_group=str_c(str_remove(pattern_id, ".[0-9]+$"), ".", name)) %>%
  mutate(pattern_group=str_remove(pattern_group, "patterns."),
         chr_str=str_c(chrom_peak, ":", start_peak, "-", end_peak)) %>%
  as_tibble()

df_distinct <- df_seqlets %>%
  distinct(chr_str, pattern_group)

ordered_levels <- df_seqlets %>%
  distinct(pattern_group) %>%
  mutate(group=if_else(str_detect(pattern_group, "^pos"),0,1),
         rank=as.integer(str_replace(pattern_group, "^.*_(\\d+)\\..*$", "\\1"))) %>%
  arrange(group, rank) %>%
  pull(pattern_group)

# df_pairs <- df_seqlets %>%
#   select(pattern_group, chr_str) %>%
#   group_by(chr_str) %>%
#   filter(dplyr::n() > 1) %>%
#   summarise(pairs=list(combn(pattern_group, 2, simplify=FALSE)), .groups='drop') %>%
#   unnest(pairs) %>%
#   mutate(pair=map_chr(pairs, ~ paste(sort(.x), collapse="-")))
# 
# df_pairs %<>%
#   count(pair, name="n_count") %>%
#   mutate(idx1=fct(str_extract(pair, "^[^-]+"), levels=ordered_levels),
#          idx2=fct(str_extract(pair, "[^-]+$"), levels=ordered_levels))
# 
# df_pairs %>%
#   pivot_wider(id_cols="idx1", names_from="idx2", values_from="n_count", names_expand=TRUE, values_fill=0) %>%
#   arrange(idx1) %>%
#   column_to_rownames("idx1") %>%
#   as.matrix()

mat_tff_tff <- matrix(0, nrow=length(ordered_levels), length(ordered_levels), dimnames=list(ordered_levels, ordered_levels))
for (i in ordered_levels) {
  n_cond <- df_distinct %>%
    group_by(chr_str) %>%
    filter(any(i %in% pattern_group)) %>%
    ungroup() %>%
    mutate(pattern_group=fct(pattern_group, ordered_levels)) %>%
    count(pattern_group, .drop=FALSE) %>%
    pull(n, pattern_group)
  
  mat_tff_tff[i,] <- n_cond[ordered_levels]
}
```

### Total Counts (column TFFs found in peaks with row TFF)
```{r fig.width=12, fig.height=12}
pheatmap(mat_tff_tff, cluster_rows=FALSE, cluster_cols=FALSE)
```


### Fraction of Co-occurences (column TFFs found in peaks with row TFF)
```{r plt_heat_cooccur_frac_hits, fig.width=12, fig.height=12}
frac_tff_tff <- mat_tff_tff / diag(mat_tff_tff)
pheatmap(frac_tff_tff, cluster_rows=FALSE, cluster_cols=FALSE)

pheatmap(frac_tff_tff, cluster_rows=TRUE, cluster_cols=TRUE)
```


```{r plt_heat_cooccur_frac_hits_na, fig.width=12, fig.height=12}
diag(frac_tff_tff) <- NA
pheatmap(frac_tff_tff, cluster_rows=FALSE, cluster_cols=FALSE)

pheatmap(frac_tff_tff, cluster_rows=TRUE, cluster_cols=TRUE)
```


### Fraction of All Peaks (column TFFs found in peaks with row TFF)
```{r plt_heat_cooccur_frac_all, fig.width=12, fig.height=12}
pheatmap(mat_tff_tff/n_total_raw, cluster_rows=FALSE, cluster_cols=FALSE)
```


### Significantly-Enriched Co-occurances
```{r compute_tests}
df_tests <- mat_tff_tff %>%
  as_tibble(rownames="tff_cond") %>%
  pivot_longer(cols=-1, names_to="tff_hit", values_to="freq_both") %>%
  mutate(tff_cond=fct(tff_cond, ordered_levels),
         tff_hit=fct(tff_hit, ordered_levels)) %>%
  mutate(total_cond=diag(mat_tff_tff)[tff_cond],
         total_hit=diag(mat_tff_tff)[tff_hit],
         total_all=n_total_raw) %>%
  mutate(freq_cond=total_cond-freq_both,
         freq_hit=total_hit-freq_both,
         freq_none=total_all - (freq_both + freq_cond + freq_hit)) %>%
  mutate(across(starts_with("freq_"), ~ .x / total_all, .names="frac_{.col}")) %>%
  filter(as.integer(tff_hit) > as.integer(tff_cond)) %>%
  rowwise() %>%
  mutate(
    test=list(fisher.test(matrix(c(freq_none,freq_cond,freq_hit,freq_both), nrow=2))),
    tidy_test=list(tidy(test))
  ) %>% unnest(tidy_test) %>% select(-test) %>%
  mutate(q_val=p.adjust(p.value, "BH"))

df_tests %>%
  filter(q_val < 0.05, freq_both > 10, estimate > 2) %>%
  arrange(-estimate)
```

## Co-occurrence conditional fraction
Bars here show the fraction of peaks with the row TFF that also have the column TFF.
```{r plot_conditional_fraction, fig.width=8, fig.height=20}
df_tests %>%
  filter(q_val < 0.01, freq_both > 10, estimate > 2) %>%
  mutate(tff_cond_fmt=fct_relabel(tff_cond, ~ str_replace(.x, fixed("."), "\n"))) %>%
  ggplot(aes(x=tff_hit, y=freq_both/total_cond)) +
  geom_bar(stat='identity', fill="steelblue", color="black", linewidth=0.2, width=0.8) +
  facet_grid(rows=vars(tff_cond_fmt)) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0.01,0.2,0)) +
  labs(x=NULL, y="Fraction of Conditional Co-occurrence") +
  theme(strip.text = element_text(size=6))
```

## Outliers without Fewer Overlaps

```{r tbl_rarer, fig.width=8, fig.height=20}
df_tests %>%
  filter(q_val < 0.01, estimate < 0.2)
```

```{r plt_all_estimates, fig.width=4, fig.height=8}
df_tests %>%
  pivot_longer(cols=starts_with("tff"), names_to="ttf_group", values_to="pattern_group") %>%
  filter(estimate > 0) %>%
  mutate(pattern_group_ord=fct_reorder(pattern_group, estimate)) %>%
  ggplot(aes(x=log10(estimate), y=pattern_group_ord)) +
  geom_vline(xintercept=0, color="orchid", linetype="dashed") +
  geom_boxplot(outliers=FALSE) +
  geom_jitter(width=0, height=0.2, size=0.5, alpha=0.3) +
  coord_cartesian(xlim=c(-2,2)) +
  labs(x="Log10(Odds Ratio)", y=NULL)
```

---

# Runtime Details
## Session Info
<details>
```{r sesh_info, echo=FALSE}
sessionInfo()
```
</details>

## Conda Environment
<details>
```{bash conda_info, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  PREFIX=$(dirname $(dirname $(which R)))
  conda env export -p $PREFIX
fi
```
</details>
